---
title: "SDSC"
author: "Dhruv"
date: "2026-05-07"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r}
# ---------------------------
# 1. Load Libraries
# ---------------------------

library(tidyverse)
library(janitor)
library(skimr)
library(janitor)
#install.packages("skimr")
library(ggplot2)
library(plotly)
library(pROC)
library(car)
library(dplyr)
library(lme4) 
library(tidyr)
library(corrplot)
library(broom)   

```



# Data Importing  
```{r setup, include=FALSE}
# dynamically configures the input directory via an interactive file explorer. 
# eliminates the need for hardcoded file paths or technical code adjustments.

my_file_path <- file.choose("Copy of ds_take_home_clinical_outcomes.csv - Sheet1.csv")
clinical <- read.csv(my_file_path)



my_file_path_pro <- file.choose("ds_take_home_procedure_metrics.csv - Sheet1.csv")
procedure <- read.csv(my_file_path_pro)
```


# 1st data checking
```{r}

# View(clinical)
nrow(clinical) # 120 rows found
ncol(clinical) # 5 column in the data 
summary(clinical)
glimpse(clinical)
head(clinical,2)

# View(procedure)
nrow(procedure)# 300 rows found 
ncol(procedure)# 12 column found
summary(procedure)
glimpse(procedure)
head(procedure,2)

# merging the entire data 
outcome_matrix <- left_join(procedure, clinical, by = "procedure_id")
#View(outcome_matrix)
nrow(outcome_matrix)# 300 rows found after merging the data set
ncol(outcome_matrix)# 16 column (LoL, as expected)
head(outcome_matrix,10)
```





# data cleaning 
```{r}
summary(outcome_matrix)
glimpse(outcome_matrix)

outcome_matrix <- clean_names(outcome_matrix)

# na values checking 
colSums(is.na(outcome_matrix)) # Count NAs per column
mean_na <- mean(is.na(outcome_matrix)) # Calculate total proportion
mean_na # overall we 15.7% NA in data (ITS STILL IR RELEVENT) (as we need all the data) 
# checking duplicate values 
outcome_matrix[duplicated(outcome_matrix), ]
# no duplicate found in data set


```



## Correlation Heatmap
```{r}


metric_col <- select(outcome_matrix,
                     path_efficiency,
                     movement_smoothness,
                     idle_time_pct,
                     tool_switching_per_min,
                     site_visibility_pct,
                     two_tools_pct,
                     speed,
                     workspace_coverage,
                     composite_score)

M <- cor(metric_col, use = "complete.obs")
corrplot(M, method = "color")
corrplot(M, method = "color", type = "upper", addCoef.col = "black")

# The composites score mostly correlates the strongly path efficiency which is.6 percent and the second best will be Site visibility percentage which is .37.
#There is negative correlations with composite and tool switch per minute time which is negative oh 0.47. They are inversely proportional to each other If one increases the other Decreases definitely.

#One can also save this correlation plot there is no strong very strong correlations with each other Variables.

# (more idle time -> worse scores)
```



## Over/Under‑weighted Metrics
```{r}
metric_cols <- c("path_efficiency", "movement_smoothness", "idle_time_pct",
                 "tool_switching_per_min", "site_visibility_pct", "two_tools_pct",
                 "speed", "workspace_coverage")


# standardise each metric (mean 0, sd 1)
df_std <- outcome_matrix %>%
#  mutate(across(all_of(c(metric_cols, "composite_score")), scale))
  mutate(across(all_of(metric_cols), scale))

colSums(is.na(df_std)) # Count NAs per column
mean_na <- mean(is.na(outcome_matrix)) # Calculate total proportion
mean_na # 15.7 na found in data set. 
```




# Regression on standardized variables

```{r}

df_std <- outcome_matrix %>%
  mutate(across(all_of(metric_cols), scale))


df_norm <- lm(data = df_std[, metric_cols],df_std$composite_score ~ 0 + .)
summary(df_norm)


coef_df <- tidy(df_norm)
ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Estimated Metric Weights",
    x = "Metric",
    y = "Regression Coefficient"
  ) +
  theme_minimal()


# Path Efficiency, two tool percentage, site visibility percentage Are the three domnates the compiste variable. 

# On the same speed and tool switching per minute is negative weight on composite variable. This implies that higher the value for this lower the composite score drops.
```



```{r}
#----------------------------
# Association with Clinical Outcomes
#----------------------------

df_out <- outcome_matrix %>% 
  filter(!is.na(complication))

logit_comp <- glm(complication ~ composite_score, data = df_out, family = binomial)
summary(logit_comp)

# 1 unit increase in the composite result in increase of complication by 0 .09 . 
# Because this value is positive high schoolers are associated with higher risk of complications

# The P value is 0.1055 which means it is statistically not significant this means there is not even evidence in the data set to confidently say that the composite score influence the risk of complication

# Odds ratio for 10‑point increase
exp(coef(logit_comp)[2] * 10)

# While the 10-point increase in score suggests a clinically relevant increase in risk (2.49 times higher odds), the relationship is not statistically significant at the 5% level (p=0.1055).

# AIC (74.752): This value is used to compare models (lower is better)

# ---------------------------
# Complete resection
#----------------------------
df_res <- outcome_matrix %>% filter(!is.na(complete_resection))

logit_res <- glm(complete_resection ~ composite_score, data = df_res, family = binomial)
summary(logit_res)

#One unit increase in composite score will result in increase of OH .1056 you need increase in the complete reselection score. Always assuming the rest variables are constant

# P value is 0.07 which is still greater than 0105 which means this is not significant therefore there is not even evidence in the data set to confidently say the composite score Influence the risk of complication. 

# ---------------------------
# Hospital stay
#----------------------------

lm_hosp <- lm(hospital_stay_days ~ composite_score, data = df_out)
summary(lm_hosp)

#One unit increase in the composite score will an increase of 0.1095 unit in the hospital stays then. 
#The P value is greater than 0.05 that means it is statistically insignificant that means there is not even evidence in the data set to confidently say that the composite score influence the risk of hospital days()


# Overall all the three of have no significant linear relationship with the composite variable. 


```

# Addressing Confounding variables 
```{r}
# Logistic regression with procedure length
logit_conf <- glm(complication ~ composite_score + procedure_length_min,
                  data = df_out, family = binomial)
summary(logit_conf)


# The data still remains insignificant as the P value is greater than 0.05

# Mixed‑effects model with random intercept for surgeon (if surgeon_id exists)
# lme4 handles binary outcomes via glmer
if("surgeon_id" %in% names(df_out)) {
  mixed_logit <- glmer(complication ~ composite_score + procedure_length_min + (1 | surgeon_id),data = df_out, family = binomial)
  summary(mixed_logit)
}

# One can clearly say that the composite score and the procedure length in minutes Their P value is statistically not significant since it's less than 0.05.

# Your mixed-effects model shows a random effect variance of 0.3733. This indicates there is some variability in complication rates between surgeons, but it’s relatively modest.

```






# part 2 


## 2.1 Surgeon‑level Nesting
```{r}

# For linear model (hospital stay), use lmer
mixed_linear <- lmer(hospital_stay_days ~ composite_score + (1 | surgeon_id),data = df_out)
summary(mixed_linear)


# ICC = variance_surgeon / (variance_surgeon + residual variance)
var_comp <- as.data.frame(VarCorr(mixed_linear))
icc <- var_comp$vcov[1] / (var_comp$vcov[1] + var_comp$vcov[2])
cat("ICC for hospital stay:", round(icc, 3), "\n")

# Only about 6.6% of the variation in hospital stays is due to which surgeon performed the procedure. The remaining 93.4% is due to individual patient differences or random factors.

# The model suggests that neither the composite score nor the specific surgeon has a massive impact on the length of the hospital stay in this specific dataset.



```


## 2.2 Most Predictive Individual Metrics

```{r}

# Standardise metrics for coefficient comparison
df_scaled <- df_out %>%
  mutate(across(all_of(metric_cols), scale)) %>%
  mutate(across(c(procedure_length_min), scale))

logit_full <- glm(complication ~ . - procedure_id - surgeon_id - composite_score - complication_severity - complete_resection - hospital_stay_days,
                  data = df_scaled, family = binomial)
summary(logit_full)

# Compare AUC of composite vs full model
roc_comp <- roc(df_out$complication, df_out$composite_score)




used_rows <- as.numeric(rownames(model.frame(logit_full)))
pred_full <- predict(logit_full, type = "response")
roc_full <- roc(df_out$complication[used_rows], pred_full)

auc(roc_comp)
auc(roc_full)


#The full model (AUC ≈ 0.78) performs much better than the composite alone (AUC ≈ 0.66).
# The composite misses important information contained in the individual metrics.

```



# Visual's
## Composite Score Distribution
```{r}


# Create a ggplot histogram
p <- ggplot(outcome_matrix, aes(x = composite_score)) + 
  geom_histogram(bins = 30, fill = "blue", color = "white")+
   labs(title = "Distribution of composite surgical scores", x = "Composite score", y = "Count")

# Convert to interactive plotly object
ggplotly(p)

# Most procedures score between 55 and 75; the distribution is roughly symmetric.

```
Insight:
Are most procedures high-performing?( Yes looks like it most of the procedures high-performing ) 
Is there a binomial distribution? ( almost binomial distribution)
is this skewed ? ) (Nah) 

## Complication vs Composite Score
```{r}

box_plot <- plot_ly(df_out, x = ~complication, y = ~composite_score, type = "box")%>%
  layout(title = "Composite score by complication status",
         xaxis = list(title = "Complication (0 = no, 1 = yes)"),
         yaxis = list(title = "Composite score"))
box_plot

```

alright (Most important clinical insight)
Now we can when there is no complication the median value is 64.6 for a composite And the Max it goes 79.6 The composite score which is not bad. The box PL like a bell shaped curve.
However when there is some complication found it looks like it's skewed to the left and even though in that situation the low composite score is still higher when there is no complication which is 57 but the Max value is 75 composite score. One can play very safe with zero complication but still with the complication the composite score is quite not bad. 

Even though the difference is not statistically significant, the plot shows slightly higher median scores in the complication group – counter‑intuitive


## Surgeon Performance Comparison
```{r}

# Sample Data Preparation
df <- outcome_matrix
surgeon_data <- df %>%
  group_by(surgeon_id) %>%
  summarise(avg_comp_score = mean(composite_score),
     avg_complications = mean(complication),avg_duration = mean(procedure_length_min)) %>%
  pivot_longer(cols = -surgeon_id, names_to = "metric", values_to = "value")

# Visualization
ggp <- ggplot(surgeon_data, aes(x = surgeon_id, y = value, fill = surgeon_id)) +
  geom_bar(stat = "identity") +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal() +
  labs(title = "Surgeon Performance Comparison",
       y = "Average Value",
       x = "Surgeon ID") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplotly(ggp)

```

## Complication Severity Breakdown
```{r}

plot_data <- outcome_matrix %>%
  group_by(surgeon_id,complication_severity) %>%
  summarise(count = n())

fig <- plot_ly(plot_data, x = ~surgeon_id, y = ~count, type = 'bar', 
               name = ~complication_severity, color = ~complication_severity) %>%
  layout(barmode = 'stack',
         title = 'Complication Severity Breakdown by Clinical Outcome',
         xaxis = list(title = 'Clinical Outcome'),
         yaxis = list(title = 'Number of Patients'))

fig


```

One can clearly see top 10 surgeons here and they have plotted on the Y axis that the number of patients they have handled over the time span.
One can clearly say surgeon 7 has the highest number of patients lowest complication including mild 1. 
One can also say that the surgeon 2 is not a good surgeon because he had many complication over his time with mild moderate.
Surgeon 3 has the one and only one Severe complication And he has turn on overall 8 patients. 

## Hospital Stay vs Composite Score
```{r}
#Shows operational efficiency impacts.
# Visualization: Scatter plot

# Using the built-in 'mtcars' dataset
pp <- ggplot(outcome_matrix, aes(x = hospital_stay_days, y = composite_score,color = complication_severity)) + 
  geom_point()
ggplotly(pp)

# Better surgical metrics → shorter stays
```

One plane clearly see from the scatter plot that if the number of hospital days increases for a patient the composite score also tends to decrease On the same time one can also say the number of complexities severity higher is the number of stays in days is higher.

# ---------------------------
# 6. Final Linear Model:
# ---------------------------
```{r}

model1 <- lm(composite_score ~ path_efficiency +
               movement_smoothness +
               idle_time_pct +
               tool_switching_per_min +
               site_visibility_pct +
               two_tools_pct +
               speed +
               workspace_coverage,
             data = outcome_matrix )

summary(model1)
## 
## equation: 33.9 + 17.17*path_eff + 10.17*movement_smooth -0.36* idle_time_pct
##              -0.05*tool_swtic +0.19*site_visi + 0.13*two_tools - 3.07*speed
##              + 0.08*wrokspace_coverage

# Interpretation for speed. 
# Every unit increase speed the composite score will be decreasing by 3.07 units.
# That means higher the speed goes lower the composite score is gonna go.

# One can also say that all the P values less than 0.05 that means it is statistically significant to keep all the variables. 

# This also Implies that we reject the null hypothesis and accept the alternate hypothesis that means there some between the composite variables and the other variables.


# Standardized coefficients
library(lm.beta)
lm.beta(model1)


# Path Efficiency (0.385): This is your strongest positive predictor. For every 1 SD increase in efficiency, the outcome increases by ~0.39 SD.
# Tool Switching per Min (-0.339): This is your strongest negative predictor. High rates of tool switching significantly lower the outcome.
# Speed (-0.273): Interestingly, higher speed is associated with a decrease in your outcome.

# Multicollinearity check
vif(model1)

# VIF = 1: Not correlated.1 < VIF < 5: Moderately correlated.
# sooooooooooo not much of correlation here

```

## t-test
```{r}
outcome_matrix %>%
  group_by(complication) %>%
  summarise(
    mean_score = mean(composite_score, na.rm = TRUE),
    sd_score = sd(composite_score, na.rm = TRUE),
    n = n()
  )

#complication   mean_score 	sd_score 	n
#  0	      	  64.48716	5.920024	109		
#  1	  	      67.59091	6.055321	11	
#  NA	    	    64.58444	5.750626	180	

# All the mean values for the complication looks close enough Even the standard deviation is close enough for the no complication for complication or for nothing. 
# However the sample varies a lot for no complication and for yes complication But this is a real data so we cannot do much here.


# T-test
# Syntax: t.test(dependent_var ~ independent_var, data = my_dataframe)
t.test(composite_score ~ complication,data = outcome_matrix)

#Now based on the conclusion one can say we failed to reject the null hypothesis.
#That means there is difference you see between the groups and could likely be due to randomness change rather than the complication itself


#So the T test value is -1.6235 that does imply how many standard errors the group means are from each other a smaller value is generally suggest that this is relatively small given the spread of the data.

#Looking at the degree of freedom it suggests the size of the data set is quite small which is very difficult or hard to prove a difference is significant unless the gap between these things is huge However we are dealing with the real life data and this is how things work So assuming all I said is true.

```

## Attempt with Firth’s logistic regression (rare events)
```{r}

table(outcome_matrix$surgeon_id,
      outcome_matrix$complete_resection)
      
library(logistf)

firth_model <- logistf(complete_resection ~ composite_score + procedure_length_min + surgeon_id,data = outcome_matrix)

summary(firth_model)



# High Success Rate: You have ~71 successes and only 11 failures. When the outcome is almost always the same (success), there isn't enough "variance" for the model to find a pattern.
# I am still ignoring those 218 NaN observations. In a small dataset, losing 70% of your rows is usually fatal for statistical significance.

# Even with Firth correction, the composite score remains non‑significant
```




